---
title: "Spring 2017, Seminar 02"
author: "Christopher Prener, Ph.D."
date: "25 Jan 2017"
output:
  pdf_document: default
  html_notebook: default
---

### Data
Today, we'll use two data sources for our class. The examples we'll discuss come from [fivethirtyeight.com](https://fivethirtyeight.com). The [data](https://github.com/fivethirtyeight/data) was originally used for [this article](https://fivethirtyeight.com/datalab/do-pulitzers-help-newspapers-keep-readers/) that investigated whether [Pulitzer Prizes](http://www.pulitzer.org) helped newspapers keep readers.

```{r}
prize <- read.csv("pulitzer-circulation-data.csv", stringsAsFactors = FALSE)
```

Pay close attention to the structure of the code above. We've assigned the data in the `csv` file to an object named `prize`. We use the `read.csv()` function to do this. You *must* place the filename in quotes

Now, practice writing this code on the other dataset in the seminar's directory: `auto2016.csv`. These data come from the [U.S. Department of Energy](https://www.fueleconomy.gov/feg/download.shtml) and have been extensively cleaned by Chris. Name your dataframe `auto`:

```{r}

```

### Dependencies

We'll need the `psych` package for calculating descriptive statistics. You'll need to have installed this (and the `dplyr` and `ggplot2` packages) before executing this code. You can use the `install.packages()` function to install these if you have not already done so.

```{r}
library(psych)
```

We will also need the package `ggplot2`, which is used for creating plots, and the package `dplyr`, which is used for managing data:

```{r}
library(ggplot2)
library(dplyr)
```

Finally, we'll need the package `nortest` for assessing the normality of our data:

```{r}
library(nortest)
```


### Descriptive Statistics

Base `R` comes with a simple `summary()` function for obtaining descriptive statistics:

```{r}
summary(prize)
```

By adding a variable name, you can limit the output to a single variable. Remember that `R` uses the `$` symbol to indicate a variable that is stored as part of a data frame.

```{r}
summary(prize$Newspaper)
```

Now you should obtain descriptive statistics for the entire `auto` dataframe:

```{r}

```

Try obtaining descriptive statistics only for the variable `mpgHwy`, which measures the highway efficiency of a car's engine:

```{r}

```

### Renaming Variables

One thing you will hopefully notice quickly is that the variable names are much tidier in the `auto2016.csv` data than they are in the Pulitzer Prize data. We can use the `dplyr` package's `rename()` function to rename those variables:

```{r}
prize <- rename(prize, winFinal14 = Pulitzer.Prize.Winners.and.Finalists..2004.2014)
```

Now write code to rename the rest of the variables in the Pulitzer Prize data:
*  `Newspaper` = `name`
*  `Daily.Circulation..2004` = `circ2004`
*  `Daily.Circulation..2013` = `circ2013`
*  `Change.in.Daily.Circulation..2004.2013` = `circChange`
*  `Pulitzer.Prize.Winners.and.Finalists..1990.2003` = `winFinal03`
*  `Pulitzer.Prize.Winners.and.Finalists..1990.2014` = `winFinalTot`

```{r}

```

If you execute the `summary()` function again on the Pulitzer Prize data, you'll see the renaming the variables has increased the readability of the output:

```{r}

```

### More Descriptive Statistics
The `psych` package offers an alternative to the `summary()` function that includes a wider array of descriptive statistics. This package somes with a function called `describe()` that contains some of the same descriptive statistics as the `summary()` function. However, one of the advantages of the `psych` package is that it will also report skew and kurtosis as well.

```{r}
describe(prize)
```

Now try using the `psych` package to generate descriptive statistics for the `mpgHwy` variable in the `auto` dataframe:

```{r}

```

### Discriptive Plots
There are two plots that we routinely use to evaluate the distribution of our variables - histograms and bar plots.

#### Histograms

The basic `ggplot` call requires the assignment of a dataframe and an asthetic variable in the first line to an object named `winFinal14`. The second line calls the specific type of plot we want to make as a "geom". In this case, we'll use the `geom_histogram()`. The `binwidth = val` can be adapted as needed to adjust the appearance of the plot. Note that we call the object `winFinal14` afterwards to display it.

```{r}
winFinal14 <- ggplot(prize, aes(x=winFinal14)) +
  geom_histogram(binwidth = 5)
winFinal14
```

Now try creating a histogram for the `mpgHwy` variable in the `auto` dataframe:

```{r}

```

#### Bar Charts

In this example, we assign our plot of the `auto` dataframes's `manual` variable to the `manual` object. We use the `geom_bar()` function to create a bar plot of these data, which indicate whether or not a vehicle has a manual transmission.

```{r}
manual <- ggplot(auto, aes(x=manual)) +
  geom_bar()
manual
```

Now try creating a bar plot of the `auto` dataframes's `drive` variable, which contains data on whether a vehicle is two wheel, four wheel, or all wheel drive:

```{r}

```

#### Saving Plots
We can use the `ggplot2` function `ggsave()` to save our plots:

```{r}
ggsave(file="winFinal14.png")
```

Since we are using a project, all of our data will be automatically saved into the project directory.

Note that we save it as `png` file, which is a good choice for images to be placed in slides or other digital documents. Another common file format is as a `pdf` file.

Now try writing code to export the other plots you've created:

```{r}

```

### Normality Testing
One of the initial tasks for data analysis is ensuring that your dependent variable is normally distributed. Skew and kurtosis both serve as indicators of normality, but there are other ways we can assess normality.

#### Hypothesis Testing
We can use two common hypothesis tests - the Shapiro-Wilk and Shapiro-Francia tests. Shapiro-Wilk is valid for datasets with $3 \leq n < 3000$ observations. The Shapiro-Francia testis valid for datasets with $5 \leq n < 5000$ observations. 

As an aside, we should point out that text placed between dollar signs in RNotebooks is treated as LaTeX syntax. This is an excellent choice for embedding mathmatic notation and formulas in your output.

The Shapir-Wilk test uses the function `shapiro.test()` from the `stats` package and the Shair-Francia test uses the function `sf.test()` from the `nortest` package. They both have the same null hypothesis, which is that the variable in question is normally distributed. Each test's alternative hypothesis is that the data are *not* normally distributed.

```{r}
shapiro.test(prize$winFinal14)
sf.test(prize$winFinal14)
```

In both cases, our test results are statistically significant (as evidenced by p-values that need scientific notiation to be displayed!). We therefore reject the null and take on our alternative hypothesis that the variable is *not* normally distributed.

Now try assessing the normality of the `mpgHwy` variable in the `auto` dataframe:

```{r}

```

#### Q-Q Plot Using `nortest`

We can use a quantile plot to confirm the results of our hypothesis tests. The quantile plot, or q-q plot for short, plots our given distribution (represented by the dots) against the normal distribution (represented by the line). 

```{r}
winFinal14_QQ <- qqnorm(prize$winFinal14); qqline(prize$winFinal14)
```

Now try assessing the normality of the `mpgHwy` variable in the `auto` dataframe:

```{r}

```

#### Q-Q Plot using `ggplot2`

Creating a q-q plot in `ggplot2` takes a bit more effort. You need to actually estimate the parameters that a q-q plot is based on:

*  The parameter `y` is the range between the 25th percentile and the 75th percentile, also known as the interquartile range or IQR. 
*  The parameter `x` contains the IQR of the normal distribution. 
*  The parameter `slope` is the slope of the line on the plot
*  The parameter `int` is the y-intercept of the line

```{r}
y     <- quantile(prize$winFinal14, c(0.25, 0.75))
x     <- qnorm(c(0.25, 0.75))
slope <- diff(y) / diff(x)
int   <- y[1] - slope * x[1]
```

Once you have the parameters estimated, you can plot them using `ggplot2`:

```{r}
winFinal14_QQ2 <- ggplot(prize, aes(sample=winFinal14)) + 
  stat_qq(distribution=qnorm) + 
  geom_abline(intercept=int, slope=slope)
winFinal14_QQ2
```
