---
title: "Spring 2017, Seminar 02"
author: "Christopher Prener, Ph.D."
date: "25 Jan 2017"
output:
  pdf_document: default
  html_notebook: default
---

### Data
Today, we'll use two data sources for our class. The examples we'll discuss come from [fivethirtyeight.com](https://fivethirtyeight.com). The [data](https://github.com/fivethirtyeight/data) was originally used for [this article](https://fivethirtyeight.com/datalab/do-pulitzers-help-newspapers-keep-readers/) that investigated whether [Pulitzer Prizes](http://www.pulitzer.org) helped newspapers keep readers.

```{r}
prize <- read.csv("pulitzer-circulation-data.csv", stringsAsFactors = FALSE)
```

Pay close attention to the structure of the code above. We've assigned the data in the `csv` file to an object named `prize`. We use the `read.csv()` function to do this. You *must* place the filename in quotes

Now, practice writing this code on the other dataset in the seminar's directory: `auto2016.csv`. These data come from the [U.S. Department of Energy](https://www.fueleconomy.gov/feg/download.shtml) and have been extensively cleaned by Chris. Name your dataframe `auto`:

```{r}

```

### Dependencies

We'll need the `psych` package for calculating descriptive statistics. You'll need to have installed this (and the `dplyr` and `ggplot2` packages) before executing this code. You can use the `install.packages()` function to install these if you have not already done so.

```{r}
library(psych)
```

We will also need the package `ggplot2`, which is used for creating plots, and the package `dplyr`, which is used for managing data:

```{r}

```

Finally, we'll need the package `nortest` for assessing the normality of our data:

```{r}

```


### Descriptive Statistics

Base `R` comes with a simple `summary()` function for obtaining descriptive statistics:

```{r}
summary(prize)
```

By adding a variable name, you can limit the output to a single variable. Remember that `R` uses the `$` symbol to indicate a variable that is stored as part of a data frame.

```{r}
summary(prize$Newspaper)
```

Now you should obtain descriptive statistics for the entire `auto` dataframe:

```{r}

```

Try obtaining descriptive statistics only for the variable `mpgHwy`, which measures the highway efficiency of a car's engine:

```{r}

```

### Renaming Variables

One thing you will hopefully notice quickly is that the variable names are much tidier in the `auto2016.csv` data than they are in the Pulitzer Prize data. We can use the `dplyr` package's `rename()` function to rename those variables:

```{r}
prize <- rename(prize, winFinal14 = Pulitzer.Prize.Winners.and.Finalists..2004.2014)
```

Now write code to rename the rest of the variables in the Pulitzer Prize data:
*  `Newspaper` = `name`
*  `Daily.Circulation..2004` = `circ2004`
*  `Daily.Circulation..2013` = `circ2013`
*  `Change.in.Daily.Circulation..2004.2013` = `circChange`
*  `Pulitzer.Prize.Winners.and.Finalists..1990.2003` = `winFinal03`
*  `Pulitzer.Prize.Winners.and.Finalists..1990.2014` = `winFinalTot`

```{r}

```

If you execute the `summary()` function again on the Pulitzer Prize data, you'll see the renaming the variables has increased the readability of the output:

```{r}

```

### More Descriptive Statistics
The `psych` package offers an alternative to the `summary()` function that includes a wider array of descriptive statistics. This package somes with a function called `describe()` that contains some of the same descriptive statistics as the `summary()` function. However, one of the advantages of the `psych` package is that it will also report skew and kurtosis as well.

```{r}
describe(prize)
```

Now try using the `psych` package to generate descriptive statistics for the `mpgHwy` variable in the `auto` dataframe:

```{r}

```
